---
title: "Flexdashboard SCUC Data Analysis"
author: "Dan Swart"
output: 
  flexdashboard::flex_dashboard:
    css: swart-20250327.css
    orientation: columns
    vertical_layout: fill
    includes:
      in_header: "header.html"
runtime: shiny
---


```{css css-styling-download-buttons, echo=FALSE}
/* Styling for download buttons */
.btn-default {
  background-color: #2E86C1;
  color: white;
  border-color: #1A5276;
  margin-bottom: 10px;
  transition: all 0.3s;
}

.btn-default:hover {
  background-color: #1A5276;
  color: white;
}

/* Add a floating download section to each chart */
.download-section {
  position: absolute;
  top: 10px;
  right: 10px;
  z-index: 1000;
  background-color: rgba(255, 255, 255, 0.8);
  padding: 5px;
  border-radius: 5px;
  box-shadow: 0 0 5px rgba(0,0,0,0.2);
}

```


```{css css-styling-plot-containers, echo=FALSE}

/* Create a container with a padding-bottom of 40% (which gives the 10:4 ratio) */

<!-- .chart-wrapper { -->
<!--   display: flex; -->
<!--   justify-content: center; -->
<!--   align-items: center; -->
<!--   width: 100%; -->
<!--   height: 100%; -->
<!-- } -->

<!-- <!-- .aspect-ratio-container { --> -->
<!-- <!--   width: 100%; --> -->
<!-- <!--   height: 100%; --> -->
<!-- <!--   position: relative; --> -->
<!-- <!--   padding-bottom: 40%; /* This creates the 10:4 ratio (40% = 4/10 * 100%) */ --> -->
<!-- <!--   margin-bottom: 50px; /* Add some space below */ --> -->
<!-- <!-- } --> -->


<!-- .aspect-ratio-container { -->
<!--   width: 100%; -->
<!--   height: 100%; -->
<!--   position: absolute; -->
<!--   top: 0; -->
<!--   left: 0; -->
<!--   padding-bottom: 40%; /* This creates the 10:4 ratio (40% = 4/10 * 100%) */ -->
<!--   margin-bottom: 50px; /* Add some space below * -->
<!-- } -->

/*End container styling */

```


```{r global-chunk, include=FALSE}


# Load Libraries
library(tidyverse)
library(qicharts2)
library(haven)
library(readxl)
library(skimr)
library(visdat)
library(gghighlight)
library(dplyr)
library(stringr)
library(ggtext)
library(gganimate)
library(ggplot2)
library(purrr)
library(scales)
library(ggrepel)
library(glue)             # concatenating strings
library(plotly)
library(htmlwidgets)
library(readr)
library(camcorder)
library(RColorBrewer)
library(rlang)
library(cowplot)
library(shiny)
library(flexdashboard)
library(DT)
library(htmltools)
library(shinyobjects)     # debugging shiny code
library(prettycode)
library(vroom)            # fast loading of data frames
library(tidylog)          # reports changes in console
library(janitor)
library(lubridate)
library(tsibble)          # working with time series data
library(tidytext)         # text analysis for chr columns
library(dlookr)           # diagnose and repair data frames; can compare distribs to normal
library(flextable)
library(patchwork)        # display multiple ggplot2 objects with various configs
library(shinycssloaders)
library(shinyjs)
library(shinyBS)
library(showtext)
library(see)
library(ggokabeito)
library(here)


# source_fonts <- fonts[grepl("Cabin", fonts$family, ignore.case = TRUE), ]
# print(source_fonts[, c("family", "path")])


# Add fonts directly to runtime: shiny environment
font_add("Roboto", "/Users/D/Library/Fonts/Roboto-Regular.ttf")
showtext_auto()

font_add("Cabin", "/Users/D/Library/Fonts/SourceSans3-Regular.ttf")
showtext_auto()


# Reactive data storage
data_reactive <- reactiveVal(NULL)   # stores and updates a dataset during the runtime of the app.  This object is updated by 'observeEvent(input$file_chosen, {' in the server section.

column_choices <- reactiveVal(NULL)  # stores and updates column names for selection during the runtime of the app.  This object is updated by  'observeEvent(input$file_chosen, {' in the server section

column_data <- reactiveVal(NULL)     # stores and updates row contents of selected column(s) during the runtime. of the app.  This object is updated by 'observeEvent(input$file_chosen, { '   in the server section.

# Options
options(scipen = 999)
options(qic.clshade = T)                      # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black')                # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "red")                # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "darkgreen")          # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti'))   # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2)   # Set upload maximum to 50 MB


# # Helper function for maintaining aspect ratio
# plotOutput_wrapper <- function(outputId, width = "100%", height = "100%") {
#   htmltools::tags$div(
#     class = "aspect-ratio-container",
#     plotOutput(outputId, width = "100%", height = "100%")
#   )
# }


```





# Inputs {.sidebar data-width=350}


```{r side-bar-inputs-no-server-needed}

# UI Components: These include both the sidebar UI and the outputs in the Vizualization Section of the Server code

#   Sidebar UI: Your inputs (defined with # Inputs {.sidebar})
#   Server-side outputs UI: The visualization containers (e.g., # Run Chart)


################  SIDEBAR DROPDOWNS THAT FEED SERVER FUNCTIONS BUT NEED NO SERVER CALCULATIONS  #####

# File upload chooser button
fileInput(inputId = "file_chosen",
          label = "Choose a CSV File",
          accept = c(".csv")) 

# Check if file has a header
checkboxInput(inputId = "header",
              label = "Header",
              value = TRUE)

# Dropdown to choose the number of rows to display in Table
selectInput(
  inputId = "rows_per_page",
  label = "Rows per Page:",
  choices = c(10, 30, 50, 100, 200),
  selected = 30   # default value
  )

# Dropdown to choose the type of file to create when exporting
selectInput(
  inputId = "export_format",
  label = "Select Export Format:",
  choices = c("SVG", "PNG"),
  selected = "PNG" # Default value
)

# Choose type of data on Y axis
selectInput(
  inputId = "y_axis_type",
  label = "Y-Axis Format Data Type:",
  choices = c("Percentage" = "percent", 
              "Rate" = "rate",
              "Count" = "count", 
              "Currency" = "currency",
              "Other" = "other"),
  selected = "percent"
)


selectInput(
  inputId = "divide_by_100", 
  label =   "Divide values by 100?", 
  choices = c("Yes", "No"), 
  selected = "No"
  )

selectInput(
  inputId = "staar_grouping",
  label = "Select Achievement Level for Titles:",
  choices = c("Approaches or Above", "Meets or Above", "Masters", "N/A"),
  selected = "Approaches or Above"
)


numericInput("cohort_start_grade", "Cohort Starting Grade:", 3, min = 3, max = 8)


numericInput("cohort_start_year", "Cohort Starting Year:", 2018, min = 2015, max = 2023)


####################  END OF SIDEBAR DROPDOWNS THAT FEED SERVER FUNCTIONS BUT NEED NO SERVER CALCULATIONS  #####



####################  SIDEBAR DROPDOWNS THAT DEPEND ON SERVER SECTION TO CALCULATE  #####


# Dynamic drop downs for selections appearing in sidebar, and require Server side calcs

# Inputs to select value for grouping
uiOutput("column_dropdown_choices") 

# Slider for selecting year range
uiOutput("years_chosen")

# Dropdowns for X axis column selection
uiOutput("x_column_selected")

# Dropdowns for Y axis column selection
uiOutput("y_column_selected")

# Dropdown for Chart Title selection
uiOutput("chart_title_selections")      

# Dropdown for Chart Subtitle column selection
uiOutput("chart_subtitle_selections")   

# Dropdown for selecting variable(s) that may be grouped (for all charts that can display more than 1 variable)
uiOutput("grouping_categories_selected")

# Debug Info {data-height=200}
verbatimTextOutput("debug_info")



####################  END OF SIDEBAR CHOICES THAT DEPEND ON SERVER SECTION TO CALCULATE  #####


####################  END OF SIDEBAR SECTION  #####


```






# Server Logic {.hidden}


```{r server-logic}

# Server Logic: This contains all your reactive expressions, render functions, etc.

#   The section is hidden by marking with # Server Logic {.hidden}



#######################   BEGIN SERVER SECTION  #####



#######################  'OBSERVE' FUNCTION UPDATES 3 GLOBAL REACTIVE OBJECTS  #####

# File upload observer creates 4 global storage objects called 'data', 'data_reactive', 'column_choices' and 'column_data'

observeEvent(input$file_chosen, {
  req(input$file_chosen)

  print(paste("Loading file:", input$file_chosen$name))
  
  # Read the CSV file
  data <- read_csv(input$file_chosen$datapath, col_names = TRUE)
  
  
  # Store data frame selected by user
  data_reactive(data)
  
  # Store column names
  column_choices(names(data))
  
  # Store unique values
  column_data(lapply(data, unique))
  
  print("Data processing complete")
})

# Add this to your server section
observeEvent(input$file_chosen, {
  req(column_data())
  
  # Set up server-side processing for the "value" column
  if ("value" %in% column_choices()) {
    updateSelectizeInput(
      session,
      "select_value",
      choices = c("All", column_data()[["value"]]),
      server = TRUE
    )
  }
}, once = TRUE)


##################  END OBSERVE FUNCTION UPDATES 4 GLOBAL REACTIVE OBJECTS  #####


##################  renderUI FUNCTIONS TO UPDATE DROPDOWN CHOICES  #####

# Generate dropdowns of column names for X axis selection
output$x_column_selected <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "x_column",
    label = "Select X-Axis Column:",
    choices = column_choices(),  # these are row names of selected columns
    selected = "year"       # column_choices()[1] # Default selection
  )
})


# Generate dropdown of column names for Y axis selection
output$y_column_selected <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "y_column",
    label = "Select Y-Axis Column:",
    choices = column_choices(),     # these are row names of selected columns
    selected = "value"              # column_choices()[2] # Default selection
  )
})


# Generate dropdown of column names for chart title selection
output$chart_title_selections <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "chart_title",
    label = "Select Categories to Include in Chart Title:",
    choices = column_choices(),
    selected = column_choices()[1],  # Default to first column
    multiple = TRUE  # Allow selecting multiple columns
  )
})



# Generate dropdown of column names for chart subtitle selection
output$chart_subtitle_selections <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "chart_subtitle",
    label = "Select Category to Include in Chart Subtitle:",
    choices = column_choices(),      # these are row names of selected columns
    selected = "year",               # column_choices()[2] # Default selection
    multiple = FALSE
  )
})


output$column_dropdown_choices <- renderUI({
  req(column_choices())
  
  column_ui <- lapply(column_choices(), function(col_name) {
    # Special handling for the "value" column
    if(col_name == "value") {
      selectizeInput(
        inputId = paste0("select_", col_name),
        label = paste("Select value for", col_name),
        choices = c("All"),  # Start with just "All"
        selected = "",
        multiple = TRUE,
        options = list(
          placeholder = 'Type to search values...',
          maxOptions = 10000,
          maxItems = 10000,
          loadThrottle = 200
        )
      )
    } else {
      # Normal handling for other columns
      selectizeInput(
        inputId = paste0("select_", col_name),
        label = paste("Select value for", col_name),
        choices = c("All", column_data()[[col_name]]),
        selected = "",
        multiple = TRUE
      )
    }
  })
  
  do.call(tagList, column_ui)
})


# Multi-line chart dropdown (for multiple category selection)
output$grouping_categories_selected <- renderUI({
  req(column_choices())
  selectizeInput(  # Using selectizeInput for better multiple selection
    inputId = "multi_var_group",
    label = "Select Categories for Multi-Variable Comparisons:",
    choices = column_choices(),
    selected = column_choices()[1],
    multiple = TRUE,  # Allow multiple selections
    options = list(plugins = list('remove_button'))
  )
})


# Generate dynamic year slider, if a "Year" column exists in the chosen data frame
output$years_chosen <- renderUI({
  req(data_reactive())
  
  # Check if there is a column named "Year" or similar
  year_col <- column_choices()[grepl("year", tolower(column_choices()))]
  
  # Retrieve min and max if 'year' column present for use in sliderInput
  if (length(year_col) > 0) {
    min_year <- min(data_reactive()[[year_col]], na.rm = TRUE)
    max_year <- max(data_reactive()[[year_col]], na.rm = TRUE)
    
  # Create slider input values for range of years
    sliderInput(
      inputId = "year_range",
      label = "Filter by Year Range:",
      min = min_year,
      max = max_year,
      value = c(min_year, max_year),
      step = 1,
      sep = ""
    )
  }
})


#############  END renderUI FUNCTIONS UPDATE DROPDOWN CHOICES  #####



#############  SERVER - reactive FUNCTIONS  #####

### The value returned by the reactive() expression will automatically update whenever its dependencies (i.e., inputs or other reactive values) change. 

#  'filtered_data' object collects the data resulting from User filtering choices. 

#  'data' object begins with the UNFILTERED data from the database chosen by the user OUTSIDE of the filtering code.  It can be called upon later as an reactive object on its own.

#  'data' is then filtered and becomes 'filtered_data'

filtered_data <- reactive({
  tryCatch({
    # Check if data is available
    req(data_reactive())
    
    # Start with the full data
    data <- data_reactive()
    
    # Apply column filters if any are selected
    for (col_name in column_choices()) {
      input_name <- paste0("select_", col_name)
      if (!is.null(input[[input_name]]) && length(input[[input_name]]) > 0) {
        if (!("All" %in% input[[input_name]])) {
          # Safely filter - check if column exists first
          if (col_name %in% names(data)) {
            data <- data[data[[col_name]] %in% input[[input_name]], , drop = FALSE]
          }
        }
      }
    }
    
    # Apply year range filter if a "Year" column exists
    if (!is.null(input$year_range)) {
      year_col <- column_choices()[grepl("year", tolower(column_choices()))]
      if (length(year_col) > 0 && year_col %in% names(data)) {
        year_filter <- data[[year_col]] >= input$year_range[1] & 
                        data[[year_col]] <= input$year_range[2]
        data <- data[year_filter, , drop = FALSE]
      }
    }
    
    # Return the filtered data, ensuring it's a data frame
    return(as.data.frame(data))
    
  }, error = function(e) {
    # This is inside the tryCatch
    print(paste("Error in filtered_data:", e$message))
    return(data.frame())  # Return empty data frame on error
  })
})



# Needed for cohort plot
cohort_data <- reactive({
  # Require the basic inputs
  req(filtered_data(), input$cohort_start_year, input$cohort_start_grade, input$year_range)
  
  # Print debugging info
  print("Cohort data function running")
  print(paste("Start year:", input$cohort_start_year))
  print(paste("Start grade:", input$cohort_start_grade))
  
  start_year <- as.numeric(input$cohort_start_year)
  start_grade <- as.numeric(input$cohort_start_grade)
  
  # Get base data
  all_data <- filtered_data()
  print(paste("Filtered data rows:", nrow(all_data)))
  
  # If no data, return empty frame
  if(nrow(all_data) == 0) {
    return(data.frame())
  }
  
  # Use the existing year range to determine how many years to track
  available_years <- seq(from = input$year_range[1], to = input$year_range[2])
  print(paste("Available years:", paste(available_years, collapse=", ")))
  
  # Calculate which years we can actually track based on start year and available range
  usable_years <- available_years[available_years >= start_year]
  print(paste("Usable years:", paste(usable_years, collapse=", ")))
  
  # Make sure we don't go beyond available data
  if(length(usable_years) == 0) {
    print("No usable years found")
    return(data.frame()) # Return empty data frame if no valid years
  }
  
  # Calculate corresponding grades (1 grade level per year)
  years_to_track <- length(usable_years)
  grades <- seq(from = start_grade, by = 1, length.out = years_to_track)
  print(paste("Tracking grades:", paste(grades, collapse=", ")))
  
  # Verify we have grade_level_code in our data
  if(!"grade_level_code" %in% names(all_data)) {
    print("Warning: grade_level_code column not found, looking for grade_level_name")
    # Try to use grade_level_name if available
    if("grade_level_name" %in% names(all_data)) {
      # Extract numbers from grade level names (e.g., "Grade 3" → 3)
      all_data$grade_level_code <- as.numeric(gsub("\\D", "", all_data$grade_level_name))
    } else {
      print("Error: Neither grade_level_code nor grade_level_name found")
      return(data.frame())
    }
  }
  
  # This is important: we need to make sure we're only including the right grade for each year
  final_result <- data.frame()
  for (i in 1:length(usable_years)) {
    current_year <- usable_years[i]
    current_grade <- grades[i]
    
    print(paste("Looking for year", current_year, "and grade", current_grade))
    
    year_data <- all_data %>%
      filter(year == current_year, grade_level_code == current_grade)
    
    print(paste("Found", nrow(year_data), "rows"))
    
    if(nrow(year_data) > 0) {
      final_result <- rbind(final_result, year_data)
    }
  }
  
  print(paste("Final result has", nrow(final_result), "rows"))
  
  # If we have no data, return empty frame
  if(nrow(final_result) == 0) {
    print("No matching data found for cohort")
    return(data.frame())
  }
  
  # Add a progression label for the x-axis
  final_result$progression <- paste0("Year ", final_result$year, " (Grade ", final_result$grade_level_code, ")")
  
  # Sort by year to ensure correct order in chart
  final_result <- final_result %>% arrange(year)
  
  return(final_result)
})


#############  END OF reactive FUNCTIONS  #####




#############  'OBSERVE' FUNCTION TO ANTICIPATE Y-AXIS DATA TYPE  #####

# Replace your existing observe function with this one
observe({
  req(filtered_data(), input$y_column)
  
  tryCatch({
    # Check if filtered_data is empty
    if (nrow(filtered_data()) == 0) {
      print("Filtered data is empty, cannot determine y-axis type")
      return()
    }
    
    # Check if column exists
    if (!input$y_column %in% names(filtered_data())) {
      print(paste("Column", input$y_column, "not found in filtered data"))
      return()
    }
    
    # Try to get the data
    y_data <- filtered_data()[[input$y_column]]
    
    # Check if data is numeric
    if (!is.numeric(y_data)) {
      print(paste("Column", input$y_column, "is not numeric"))
      updateSelectInput(session, "y_axis_type", selected = "other")
      return()
    }
    
    # Now safely proceed with numeric checks
    if (all(y_data >= 0 & y_data <= 1, na.rm = TRUE)) {
      updateSelectInput(session, "y_axis_type", selected = "percent")
    } else if (all(abs(y_data - round(y_data)) < 0.001, na.rm = TRUE)) {
      updateSelectInput(session, "y_axis_type", selected = "count")
    } else {
      updateSelectInput(session, "y_axis_type", selected = "other")
    }
    
  }, error = function(e) {
    print(paste("Error in y-axis detection:", e$message))
    # Default to other if there's an error
    updateSelectInput(session, "y_axis_type", selected = "other")
  })
})



##########  END 'OBSERVE' FUNCTION TO ANTICIPATE Y-AXIS DATA TYPE  #####





#########  BEGIN DATA TABLE LOGIC  #####


output$data_frame_chosen <- renderDT({
  # Add debugging
  print("Attempting to render data table")

  # Check if filtered_data exists
  if (is.null(filtered_data())) {
    print("filtered_data is NULL")
    return(datatable(data.frame(Message = "No data available")))
  }

  # Check data type and structure
  print(paste("Class of filtered_data:", class(filtered_data())))
  print(paste("Dimensions:", paste(dim(filtered_data()), collapse = "x")))

  # Make sure it's a data frame
  if (!is.data.frame(filtered_data())) {
    print("Converting to data frame")
    data_to_show <- as.data.frame(filtered_data())
  } else {
    data_to_show <- filtered_data()
  }

  # Try to render the table
  tryCatch({
    datatable(
      data_to_show,
      style = 'bootstrap',
      editable = 'cell',
      filter = 'top',
      caption = htmltools::tags$caption(
        style = 'caption-side: top;
                text-align: center;
                padding-bottom: 10px;
                font-family: "Cabin", Arial, sans-serif;
                font-weight: 900;
                font-size: 56;',
                'Table 2: ',
                htmltools::em('This is your data sorted and filtered per your inputs.')
      ),
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        striped = TRUE,
        spacing = "l",
        align = "lccccccr",
        digits = 1,
        width = "90%",
        pageLength = as.numeric(input$rows_per_page),
        paging = TRUE,
        ordering = TRUE,
        autoWidth = TRUE,
        scrollX = TRUE,
        scrollY = "400px",
        scrollCollapse = FALSE,
        buttons = c("copy", "csv", "excel", "pdf", "print"),
        responsive = TRUE
      ),
      class = "display compact stripe hover"
    )
  }, error = function(e) {
    print(paste("Error in datatable:", e$message))
    return(datatable(data.frame(Error = e$message)))
  })
})


#####################  END DATA TABLE LOGIC  #####




#####################  BEGIN RUN CHART LOGIC  #####


output$run_chart_plot <- renderPlot({
  req(filtered_data(), input$x_column, input$y_column, input$chart_title, input$year_range)

  print("Chart rendering started")
  
  # Extract x_data and y_data from filtered_data()
  data <- filtered_data()
  x_data <- data[[input$x_column]]
  y_data <- data[[input$y_column]]

  # Qualify the data
  req(!is.null(x_data), !is.null(y_data), length(x_data) > 0, length(y_data) > 0)

  # Calculate centerline (median)
  centerline <- median(y_data, na.rm = TRUE)

 
  
###############  BEGIN RUN CHART TITLES  #####
  
# Create a title_elements storage object to be filled by selection process below
title_elements <- c()     # placeholder

# Process each selected column
for(col_name in input$chart_title) {
  if(col_name %in% names(data)) {
    # If there's just one unique value in this column, use it
    if(length(unique(data[[col_name]])) == 1) {
      title_elements <- c(title_elements, as.character(unique(data[[col_name]])))
    }
    # If there are a few unique values (under 5), include them all
    else if(length(unique(data[[col_name]])) <= 5) {
      values <- paste(unique(data[[col_name]]), collapse=", ")
      title_elements <- c(title_elements, values)
    }
    # If there are many values, just indicate the count
    else {
      n_values <- length(unique(data[[col_name]]))
      title_elements <- c(title_elements, paste0(n_values, " Values"))
    }
  }
}


# Combine all elements with hyphens
title_content <- if (length(title_elements) > 0) {
  paste(title_elements, collapse = " - ")
} else {
  "Selected Data"  # Default if no columns selected
}

# Create the final title
title <- paste0(
  "<span style='color:darkorchid1'>",
  "Run Chart: ", 
  title_content,
  "</span>"
)

# Modify this subtitle section in your line chart code
req(input$year_range, input$staar_grouping)  # Make sure these inputs exist

subtitle <- paste0(
  "For Students Reaching <i>'",
  input$staar_grouping,
  "'</i><br>for Years ",  # Use HTML tags <i> for italics and <br> for line break
  input$year_range[1],
  " - ",
  input$year_range[2]
)

# # Now create the subtitle with staar_grouping
# # First make sure input$year_range and input$staar_grouping exist
# req(input$year_range, input$staar_grouping)
# 
# # Create the subtitle
# subtitle <- paste0(
#   "For Students Reaching ",
#   input$staar_grouping,
#   " for the Years ",
#   input$year_range[1],
#   " - ", 
#   input$year_range[2]
# )

# # create object named chart_subtitle
#   chart_subtitle <- input$year_range
# 
#   subtitle <- paste0("For the Years Ended ",
#                      chart_subtitle[1],
#                      " - ",
#                      chart_subtitle[2]
#   )

  
###############  END RUN CHART TITLES  #####
  
  
# Create the run chart
  rc <- ggplot(data,
               aes(x = .data[[input$x_column]],
                   y = .data[[input$y_column]]
                   )
               ) +
    
    # Add lines connecting the points
    geom_line(color = "grey",
              linewidth = 2.00) +
    
    # Add points
    geom_point(size = 6,
               color = "blue"
               ) +
    
    # Add the centerline (median)
    geom_hline(yintercept = centerline, 
               linetype = "dashed",
               color = "black",
               linewidth = 1
               ) +
    
    # Add an annotation for the centerline
    annotate(
      "text",
      x = max(x_data, na.rm = TRUE),   # Place the annotation at far right edge
      y = max(y_data, na.rm = TRUE),   # Place the annotation at top
      label = "(centerline is the median)",
      color = "red",
      size = 4,
      hjust = -.5,      # pos number = Align slightly BELOW the line
      vjust = -1        # neg number = Align slightly ABOVE the line
    ) +

    labs(
      title = title,
      subtitle = subtitle,
      caption = "Data Source:  The Texas Education Association",
      x = element_blank(), 
      y = element_blank()
      # x = "Academic Year-Ended",
      # y = "Values"
    )

  # Adjust Y axis  scales and labels based on data type
  if (input$y_axis_type == "percent") {
    rc <- rc + scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      name = "Percentage (%)"
    )
  } else if (input$y_axis_type == "rate") {
    rc <- rc + scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      name = "Percentage (%)"
    )
  } else if (input$y_axis_type == "count") {
    rc <- rc + scale_y_continuous(
      labels = scales::comma_format(),
      name = "Count"
    )
  } else if (input$y_axis_type == "currency") {
    rc <- rc + scale_y_continuous(
      labels = scales::dollar_format(),
      name = "Amount ($)"
    )
  } else {
    # Default - plain number format with appropriate decimal places
    rc <- rc + scale_y_continuous(
      labels = scales::number_format(accuracy = 0.01),
      name = "Value"
    )
  }
  
  # Set X axis  scales and labels 
rc <- rc + 
    scale_x_continuous(
      breaks = unique(data[[input$x_column]]),            # Show all unique years
      labels = function(x) format(x, scientific = FALSE)  # Format as regular numbers
    )
  
  
  # Continue adding theme elements
  rc <- rc + 
    # Set theme
    theme_minimal(
      base_size = 26,
      base_family = 'Cabin'
      ) +
    
    # Add grid lines 
    theme(
      aspect.ratio = 2/5,  # This will create a 5:2 aspect ratio (width:height).

      panel.grid.major = element_line(color = "gray",
                                      linewidth = 0.5
                                      ),
      panel.grid.minor = element_blank()
      ) +
    # remove legend
    theme(legend.position = "none",
          aspect.ratio = 2/5) +         # This will create a 5:2 aspect ratio (width:height)
    theme(
      plot.title.position = "panel",
      text = element_text(color = "royalblue"),
      plot.title = ggtext::element_markdown(
         color = "darkgreen",
         size = rel(1.25),
         face = "bold",
         lineheight = 1.1,
         margin = margin(0, 0, 0, 0, "pt")
         ),

      plot.subtitle = ggtext::element_markdown(
         color = "darkgreen",
         size = rel(1.10),
         face = "bold",
         lineheight = 0.9,
         margin = margin(5, 0, 0, 0, "pt")
         ),

      plot.caption = element_text(color = "darkgreen",
                                  size = rel(1.0),
                                  face = "bold",
                                  lineheight = 1.0,
                                  margin = margin(5, 0, 0, 0, "pt")
                                  ),

      strip.text = element_text(color = "orange",
                                size = rel(1.1),
                                face = "bold",
                                margin = margin(5, 0, 0, 0, "pt")
                                ),

      # X-axis tick label styling
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        face = "bold", 
        size = rel(1.1)
      ),
      
      # X-axis title styling
      axis.title.x = element_text(
       face = "bold", 
       size = rel(1.3),
       color = "royalblue",  # Optional: specify color
       margin = margin(t = 10)
      ),
      
      # y-axis tick label styling
      axis.text.y = element_text(
        hjust = 1,
        face = "bold", 
        size = rel(1.1)
      ),
      
       # y-axis title styling
       axis.title.y = element_text(
        face = "bold", 
        size = rel(1.1),
        color = "royalblue",  # Optional: specify color
        margin = margin(t = 10, r = 10)
       ),
      
      panel.background = element_rect(fill = "white", color = NA),
      
      plot.background = element_rect(fill = "white", color = NA)
  )

    
 
  # Save the plot to reactive values
  plot_store$run_chart <- rc

  
  # Return the ggplot object directly
  return(rc)
  
  
})   


###############  END RUN CHART LOGIC  #####





###############  BEGIN LINE CHART (CAN HANDLE MULTIPLE LINES)  #####

 # Add this debugging above your renderPlot
output$debug_info <- renderPrint({
  cat("Columns in filtered data:", paste(names(filtered_data()), collapse=", "), "\n")
  cat("Number of rows:", nrow(filtered_data()), "\n")
  cat("Is y-column present?", input$y_column %in% names(filtered_data()), "\n")
  cat("Sample of data:\n")
  print(head(filtered_data()))
})
  


output$line_chart <- renderPlot({
  print("Attempting to render line chart")
  
  # Check required inputs
  req(filtered_data(), input$x_column, input$y_column, input$chart_title, input$year_range, input$multi_var_group)
  
  print(paste("X column:", input$x_column))
  print(paste("Y column:", input$y_column))
  
  # Check if we have at least one grouping variable
  req(length(input$multi_var_group) > 0)
  print(paste("Grouping variables:", paste(input$multi_var_group, collapse=", ")))

  # Extract data from filtered_data()
  data <- filtered_data()
  
  print(paste("Filtered data dimensions:", nrow(data), "x", ncol(data)))
  
  # Check for empty data
  if(nrow(data) == 0) {
    print("Filtered data is empty, cannot render chart")
    return(ggplot() + 
           annotate("text", x = 0.5, y = 0.5, label = "No data available to plot") +
           theme_void())
  }
  
  # Check if columns exist
  if(!input$x_column %in% names(data)) {
    print(paste("X column", input$x_column, "not found in data"))
    return(ggplot() + 
           annotate("text", x = 0.5, y = 0.5, label = paste("Column", input$x_column, "not found in data")) +
           theme_void())
  }
  
  if(!input$y_column %in% names(data)) {
    print(paste("Y column", input$y_column, "not found in data"))
    return(ggplot() + 
           annotate("text", x = 0.5, y = 0.5, label = paste("Column", input$y_column, "not found in data")) +
           theme_void())
  }
  
  # Check column types
  print(paste("X column type:", class(data[[input$x_column]])[1]))
  print(paste("Y column type:", class(data[[input$y_column]])[1]))
  
  
  # Safely create the combined group column
  if(length(input$multi_var_group) == 1) {
    # If only one column is selected, use it directly
    data$combined_group <- as.character(data[[input$multi_var_group]])
    print("Using single grouping variable")
  } else {
    # For multiple columns, create a combined string
    # First, check all columns exist
    all_cols_exist <- all(input$multi_var_group %in% names(data))
    if(!all_cols_exist) {
      print("Some selected group columns don't exist in the data")
      return(ggplot() + 
               annotate("text", x = 0.5, y = 0.5, 
                        label = "Error: Some selected group columns don't exist in the data") + 
               theme_void())
    }
    
    # Now combine the columns safely
    data$combined_group <- apply(data[, input$multi_var_group, drop = FALSE], 1, 
                              function(row) paste(row, collapse = " | "))
    print("Using multiple grouping variables")
  }
  
  print("Unique groups found:")
  print(unique(data$combined_group))
  
  # Create group title for subtitle
  groups_title <- paste(input$multi_var_group, collapse = ", ")
  
  # Create title elements
  title_elements <- c()
  for(col_name in input$chart_title) {
    if(col_name %in% names(data)) {
      if(length(unique(data[[col_name]])) == 1) {
        title_elements <- c(title_elements, as.character(unique(data[[col_name]])))
      } else if(length(unique(data[[col_name]])) <= 5) {
        values <- paste(unique(data[[col_name]]), collapse=", ")
        title_elements <- c(title_elements, values)
      } else {
        n_values <- length(unique(data[[col_name]]))
        title_elements <- c(title_elements, paste0(n_values, " Values"))
      }
    }
  }

  # Combine title elements
  title_content <- if (length(title_elements) > 0) {
    paste(title_elements, collapse = " - ")
  } else {
    "Selected Data"
  }

  # Create final title and subtitle
  title <- paste0(
    "<span style='color:darkorchid1'>",
    "Line Chart: ", 
    title_content,
    "</span>"
  )

 # Modify this subtitle section in your line chart code
req(input$year_range, input$staar_grouping)  # Make sure these inputs exist

subtitle <- paste0(
  "For Students Reaching <i>'",
  input$staar_grouping,
  "'</i><br>for Years ",  # Use HTML tags <i> for italics and <br> for line break
  input$year_range[1],
  " - ",
  input$year_range[2]
)

  
  print("Creating ggplot...")
  
  
# Create the base plot with common aesthetics
alc <- ggplot(data,
             aes(x = .data[[input$x_column]],
                 y = .data[[input$y_column]],
                 color = combined_group,
                 group = combined_group
                 )
             ) +
  
  # Add lines connecting the points
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  
  # Add labels at the end of each line
  geom_text(
    # Use dplyr to get the last point for each group
    data = data %>% 
      group_by(combined_group) %>% 
      arrange(.data[[input$x_column]]) %>% 
      slice_tail(n = 1) %>%
      ungroup(),
    aes(label = combined_group),
    hjust = 0,
    vjust = 0,
    family = 'Cabin',
    fontface = 'bold',
    size = 12,
    nudge_x = 0.1
  ) +
  
    # Add points
      geom_point(size = 4) +
    
    # If we have many categories, use a different color scale
    {if(length(unique(data$combined_group)) <= 9) {
        scale_color_brewer(palette = "Set1")
      } else {
        scale_color_viridis_d()
      }} +
    
    labs(
      title = title,
      subtitle = subtitle,
      caption = "\n Source: Texas Education Agency",
      x = element_blank(), 
      y = element_blank(),
      color = NULL
    )

  # Adjust Y axis scales and labels based on data type
  if (input$y_axis_type == "percent") {
    alc <- alc + scale_y_continuous(
      labels = scales::percent_format(accuracy = 1)
    )
  } else if (input$y_axis_type == "count") {
    alc <- alc + scale_y_continuous(
      labels = scales::comma_format()
    )
  } else if (input$y_axis_type == "currency") {
    alc <- alc + scale_y_continuous(
      labels = scales::dollar_format()
    )
  } else {
    alc <- alc + scale_y_continuous(
      labels = scales::number_format(accuracy = 0.01)
    )
  }
  

# Find the max x value
x_max <- max(data[[input$x_column]], na.rm = TRUE)


# Add this to your ggplot chain
alc <- alc +
  scale_x_continuous(
    breaks = unique(data[[input$x_column]]),
    labels = function(x) format(x, scientific = FALSE),
    limits = c(NA, x_max + 1)  # Keep the lower limit as is, extend upper limit
  )
 
    print("Using continuous scale for numeric x-axis")
 
  
  # Apply the same theme as before
  alc <- alc + 
    theme_minimal(
      base_size = 26,
      base_family = 'Cabin'
      ) +
    
    theme(
      aspect.ratio = 2/5,
      panel.grid.major = element_line(color = "gray", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      plot.title.position = "panel",
      text = element_text(color = "royalblue"),
      plot.title = ggtext::element_markdown(
         color = "darkgreen",
         size = rel(1.25),
         face = "bold",
         lineheight = 1.1,
         margin = margin(0, 0, 0, 0, "pt")
      ),
      plot.subtitle = ggtext::element_markdown(
         color = "darkgreen",
         size = rel(1.10),
         face = "bold",
         lineheight = 0.90,
         margin = margin(5, 0, 0, 0, "pt")
      ),
      plot.caption = element_text(
        color = "darkgreen",
        size = rel(1.0),
        face = "bold",
        lineheight = 1.0,
        margin = margin(5, 0, 0, 0, "pt")
      ),
      strip.text = element_text(
        color = "orange",
        size = rel(1.1),
        face = "bold",
        margin = margin(5, 0, 0, 0, "pt")
      ),
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        face = "bold", 
        size = rel(1.1)
      ),
      axis.title.x = element_text(
       face = "bold", 
       size = rel(1.3),
       color = "royalblue",
       margin = margin(t = 10)
      ),
      axis.text.y = element_text(
        hjust = 1,
        face = "bold", 
        size = rel(1.1)
      ),
      axis.title.y = element_text(
        face = "bold", 
        size = rel(1.1),
        color = "royalblue",
        margin = margin(t = 10, r = 10)
      ),
      legend.position = "none",
      legend.title = element_blank(),
      legend.text = element_text(size = rel(0.9)),
      legend.key.size = unit(0.8, "cm"),
      {if(length(unique(data$combined_group)) > 10) {
         legend.key.height = unit(0.8, "cm")
       }},
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
  
  print("ggplot created successfully")
  
  # Save the plot to reactive values
  plot_store$line_chart <- alc
  

  
  # Return the ggplot object
  return(alc)
})



####################  END OF LINE CHART  #####
  





####################  BEGIN BAR CHART LOGIC  #####


output$bar_chart <- renderPlot({
  req(filtered_data(), input$x_column, input$y_column, input$chart_title, 
      input$year_range, input$multi_var_group, input$y_axis_type, input$divide_by_100)
  
  # Make sure we have at least one grouping variable selected
  req(length(input$multi_var_group) > 0)
  # Extract data from filtered_data()
  data <- filtered_data()
  
  # Safely create the combined group column
  if(length(input$multi_var_group) == 1) {
    # If only one column is selected, use it directly
    data$combined_group <- as.character(data[[input$multi_var_group]])
  } else {
    # For multiple columns, create a combined string
    # First, check all columns exist
    all_cols_exist <- all(input$multi_var_group %in% names(data))
    if(!all_cols_exist) {
      return(ggplot() + 
               annotate("text", 
                        x = 0.5, 
                        y = 0.5, 
                        label = "Error: Some selected group columns don't exist in the data") + 
               theme_void())
    }
    
    # Now combine the columns safely
    data$combined_group <- apply(data[, input$multi_var_group, drop = FALSE], 1, 
                              function(row) paste(row, collapse = " | "))
  }
  
  # Create a title showing what we're comparing
  groups_title <- paste(input$multi_var_group, collapse = ", ")
  
  # Create a combined title from user-selected columns
  title_elements <- c()
  for(col_name in input$chart_title) {
    if(col_name %in% names(data)) {
      if(length(unique(data[[col_name]])) == 1) {
        title_elements <- c(title_elements, as.character(unique(data[[col_name]])))
      } else if(length(unique(data[[col_name]])) <= 5) {
        values <- paste(unique(data[[col_name]]), collapse=", ")
        title_elements <- c(title_elements, values)
      } else {
        n_values <- length(unique(data[[col_name]]))
        title_elements <- c(title_elements, paste0(n_values, " Values"))
      }
    }
  }
  
  # Set the plot title based on collected title elements
  plot_title <- "Bar Plot"
  if(length(title_elements) > 0) {
    plot_title <- paste(title_elements, collapse = " - ")
  }
  
  # Process data according to divide_by_100 input
  if(input$divide_by_100 == "Yes") {
    # Divide values by 100 (e.g., convert 75 to 0.75)
    data[[input$y_column]] <- data[[input$y_column]] / 100
  }
  
  # Prepare y-axis label
  y_axis_label <- input$y_column
  if(input$y_axis_type == "Percentage") {
    data[[input$y_column]] <- data[[input$y_column]] * 100
    y_axis_label <- "Percentage (%)"
  } else if(input$y_axis_type == "Count") {
    y_axis_label <- "Count"
  } else if(input$y_axis_type == "Currency") {
    y_axis_label <- "Amount ($)"
  }
  
  # Ensure x-axis column is treated as a factor to show all values
  data[[input$x_column]] <- as.factor(data[[input$x_column]])
  
  
  # Create the base plot
  
  p <- ggplot(data,
              aes(
                x = .data[[input$x_column]],
                y = .data[[input$y_column]], 
                fill = .data$combined_group
              )) +
    
    geom_col(position = position_dodge(width = 0.9),
             color = "black",
             fill = 'lightblue',
             alpha = 0.95,
             linewidth = 0.5) +
    
    labs(
      title = plot_title,
      x = element_blank(), 
      y = element_blank(),
      # x = input$x_column,
      # y = y_axis_label,
      fill = groups_title
    )
  
  # Add type-specific formatting
  if(input$y_axis_type == "Percentage") {
    # Hard-code the percentage formatter
    p <- p + scale_y_continuous(
      labels = function(y) paste0(format(y, nsmall = 0), "%")
    )
  } else if(input$y_axis_type == "Currency") {
    p <- p + scale_y_continuous(labels = scales::dollar)
  } else {
    p <- p + scale_y_continuous(labels = scales::comma)
  }
  
  # Force display of all x-axis labels
  p <- p + scale_x_discrete(
    breaks = levels(as.factor(data[[input$x_column]])) # Explicitly define every break
  )
  
  # Apply theming
  p <- p + theme_minimal (
    base_size = 36,
    base_family = 'Cabin'
    ) +
    
    theme(
      panel.grid.major = element_line(
        color = "gray", 
        linewidth = 0.5
        ),
      
      panel.grid.minor = element_blank(),
      
      legend.position = "none",
      
      plot.title = element_text(
        hjust = 0.5,
        size = rel(1.00), 
        face = "bold"),
      
      axis.title = element_text(
        size = rel(1.00)
        ),
      
      axis.text = element_text(size = rel(.90)),
      
      # Ensure x-axis text is angled
      axis.text.x = element_text(angle = 45, 
                                 hjust = 1, 
                                 vjust = 1
                                 ),
      
      legend.title = element_text(size = rel(.90)),
      
      legend.text = element_text(size = rel(.90))
      
    )
  
  
  # Save the plot to reactive values
  plot_store$bar_chart <- p
  
  # Return the completed plot
  return(p)
  
})



#################  END BAR CHART LOGIC  #####





################  BEGIN UNTRENDED I-CHART LOGIC  #####



output$untrended_i_chart <- renderPlot({
  # Require necessary inputs
  req(
    filtered_data(), input$x_column, input$y_column, input$chart_title,
    input$year_range, input$multi_var_group, input$y_axis_type, input$divide_by_100
  )

  # Extract data from filtered_data() like the other charts
  data <- filtered_data()

  # Make sure we have some data
  if (nrow(data) == 0) {
    return(ggplot() +
      annotate("text",
        x = 0.5,
        y = 0.5,
        label = "No data available with current filters"
      ) +
        
      theme_void())
  }

  # Process data for control limits calculation
  data_control_limits <- data

  # Process data according to divide_by_100 input
  if (input$divide_by_100 == "Yes") {
    # Divide values by 100 (e.g., convert 75 to 0.75)
    data_control_limits[[input$y_column]] <- data_control_limits[[input$y_column]] / 100
  }

  # For percentage display, convert to percentage for visualization
  if (input$y_axis_type == "Percentage") {
    data_control_limits[[input$y_column]] <- data_control_limits[[input$y_column]] * 100
  }

  ### CALCULATE UNTRENDED CL, SD AND CONTROL LIMITS ###
  emp_cl <- mean(data_control_limits[[input$y_column]], na.rm = TRUE) # Calculate empirical centerline

  emp_sd <- mean(abs(diff(data_control_limits[[input$y_column]])), na.rm = TRUE) / 1.128 # Calculate empirical standard deviation

  emp_lcl <- max(emp_cl - (3 * emp_sd), 0) # If LCL is negative, set it to zero

  emp_ucl <- emp_cl + (3 * emp_sd) # Calculate empirical upper control limit

  # Create modified dataset for chart (using data instead of df1)
  data_modified <- data

  # Process data according to divide_by_100 input
  if (input$divide_by_100 == "Yes") {
    # Divide values by 100 (e.g., convert 75 to 0.75)
    data_modified[[input$y_column]] <- data_modified[[input$y_column]] / 100
  }
  # For percentage display, convert to percentage for visualization
  if (input$y_axis_type == "Percentage") {
    data_modified[[input$y_column]] <- data_modified[[input$y_column]] * 100
  }

  # Shewhart Rule #1: violations if value is outside UCL or LCL
  sigma_signals <- data_modified[[input$y_column]] < emp_lcl | data_modified[[input$y_column]] > emp_ucl
  emp_pts_out_of_control <- sigma_signals

  # Add the 'out_of_control' column to the dataframe
  data_modified$emp_pts_out_of_control <- emp_pts_out_of_control

  # Runs analysis for changing CL color and linetype
  runs <- sign(data_modified[[input$y_column]] - emp_cl)
  runs <- runs[runs != 0]
  runs_lengths <- rle(runs)$lengths
  n_obs <- sum(runs_lengths)
  longest_run <- max(runs_lengths)
  n_runs <- length(runs_lengths)
  n_crossings <- n_runs - 1
  longest_run_max <- round(log2(n_obs) + 3)
  n_crossings_min <- qbinom(.05, n_obs - 1, 0.5)

  runs_signal <- (longest_run > longest_run_max | n_crossings < n_crossings_min)

  # ADD EMPIRICAL CL, SD, UCL, LCL AND RUNS_SIGNAL TO data_modified
  data_modified <- data_modified %>%
    dplyr::mutate(
      emp_cl = emp_cl, # add empirical centerline
      emp_sd = emp_sd, # add empirical standard deviation
      emp_lcl = emp_lcl, # add empirical lcl
      emp_ucl = emp_ucl, # add empirical upper control limit
      runs_signal = as.logical(runs_signal) # Ensure it's a column, not a constant
    )

  # Replace zeros in the value column with NA
  data_modified[[input$y_column]] <- ifelse(data_modified[[input$y_column]] == 0,
    NA,
    data_modified[[input$y_column]]
  )

  # Compute x and y positions for geom_richtext
  x_pos <- min(data_modified[[input$x_column]], na.rm = TRUE)
  y_pos <- max(data_modified[[input$y_column]], na.rm = TRUE)
  
  ###############  BEGIN UNTRENDED I-CHART TITLES  #####
  
  # Create a combined title from user-selected columns (same as in previous function)
  title_elements <- c()
  for(col_name in input$chart_title) {
    if(col_name %in% names(data_modified)) {
      if(length(unique(data_modified[[col_name]])) == 1) {
        title_elements <- c(title_elements, as.character(unique(data_modified[[col_name]])))
      } else if(length(unique(data_modified[[col_name]])) <= 5) {
        values <- paste(unique(data_modified[[col_name]]), collapse=", ")
        title_elements <- c(title_elements, values)
      } else {
        n_values <- length(unique(data_modified[[col_name]]))
        title_elements <- c(title_elements, paste0(n_values, " Values"))
      }
    }
  }

  # Combine title elements
  title_content <- if (length(title_elements) > 0) {
    paste(title_elements, collapse = " - ")
  } else {
    "Selected Data"
  }

  # Create the final title and subtitle
  i_title <- paste0(
    "<span style='color:darkorchid1'>",
    "Expectation Chart (untrended): ",
    title_content,
    "</span>"
  )

  # Modify this subtitle section in your line chart code
req(input$year_range, input$staar_grouping)  # Make sure these inputs exist

i_subtitle <- paste0(
  "For Students Reaching <i>'",
  input$staar_grouping,
  "'</i><br>for Years ",  # Use HTML tags <i> for italics and <br> for line break
  input$year_range[1],
  " - ",
  input$year_range[2]
)


  ###############  END I-CHART TITLES  #####
  
  # Define y-axis LABEL based on User input
  y_axis_label <- input$y_column

  if (input$y_axis_type == "Percentage") {
    y_axis_label <- "Percentage (%)"
  } else if (input$y_axis_type == "Count") {
    y_axis_label <- "Count"
  } else if (input$y_axis_type == "Currency") {
    y_axis_label <- "Amount ($)"
  } else if (input$y_axis_type == "Rate") {
    y_axis_label <- "Percentage (%)"
  }

  # Create caption
  caption <- "\n Source: Texas Education Agency"

  
  # Create plot object I-chart USING GGPLOT2 ONLY
  i_chart <- ggplot2::ggplot(
    data_modified,
    aes(
      x = .data[[input$x_column]],
      y = .data[[input$y_column]]
    )
  ) +
    
    # Add lines connecting the points
    ggplot2::geom_line(
      color = "darkgray",
      linewidth = 2.0
    ) +
    
    # Use 'color' to map logical TRUE/FALSE to red/black points
    ggplot2::geom_point(aes(color = factor(emp_pts_out_of_control)),
      size = 8
    ) +
    
    ggplot2::scale_color_manual(values = c(
      "TRUE" = "red",
      "FALSE" = "blue"
    )) +
    
    # CL line type depends on value of runs_signals
    geom_line(aes(
      y = emp_cl,
      linetype = factor(runs_signal)
    )) +
    
    scale_linetype_manual(values = c(
      "FALSE" = "solid",
      "TRUE" = "dashed"
    )) +
    
    # Format emp_ucl LINE
    ggplot2::geom_line(aes(y = emp_ucl),
      color = "red",
      linetype = "solid",
      linewidth = 2
    ) +
    
    # Format emp_lcl LINE
    ggplot2::geom_line(aes(y = emp_lcl),
      color = "red",
      linetype = "solid",
      linewidth = 2
    ) +
    
    # Add labels
    ggplot2::labs(
      title = i_title,
      subtitle = i_subtitle,
      caption = caption,
      x = element_blank(), 
      y = element_blank()
      # x = input$x_column,
      # y = y_axis_label
    ) 
    

# Compute the midpoint of the x-axis based on the year range
x_midpoint <- mean(c(input$year_range[1], input$year_range[2]))



# Then add this to your ggplot chain
i_chart <- i_chart +
  
  # Disclose CL calculation - centered on x-axis
  ggtext::geom_richtext(
  aes(
    label = paste0("<b><i>Highest, Lowest and Avg Expected Values",
    "<br>",
    "Are Accurate ONLY When Your Process is Stable</i></b>"),
    x = x_midpoint,  # Use midpoint of x values to center box on x axis
    y = y_pos
  ),
  size = 10,
  color = "black",
  hjust = .50,  # Set to 0 for left justification, .5 for center
  vjust = -1,
  fill = "lightblue",
  label.color = "black",
  label.size = unit(0.5, "inch"),  # Make the text outline box larger
  label.padding = unit(c(10, 10, 5, 10), "pt"),  # top, right, bottom, left padding
  lineheight = 0.8  # Reduce the line height to bring lines closer together
)

# # Disclose CL line type - centered on x-axis
# ggtext::geom_richtext(
#   aes(
#     label = paste0("<b><i>If run rules violated, the centerline will be dashed, not solid</i></b>"),
#     x = x_midpoint,  # Use the same x_midpoint calculated earlier
#     y = min(data_modified[[input$y_column]], na.rm = TRUE)
#   ),
#   size = 10,
#   color = "black",
#   hjust = 0.5,  # Center the text horizontally
#   vjust = 3,    # Position it below the min value
#   fill = "pink",
#   label.color = "black"
# )

  # Configure x-axis to show all years
  if (is.numeric(data_modified[[input$x_column]])) {
    i_chart <- i_chart +
      ggplot2::scale_x_continuous(
        breaks = unique(data_modified[[input$x_column]]),
        expand = c(
          0.15,
          0.00
        )
      )
  } else {
    # For character/factor columns, use discrete scale
    i_chart <- i_chart +
      ggplot2::scale_x_discrete(
        expand = c(
          0.15,
          0.00
        )
      )
  }

# Configure y-axis based on chart type
if (input$y_axis_type == "percent") {
  i_chart <- i_chart +
    ggplot2::scale_y_continuous(
      labels = function(y) paste0(format(y, nsmall = 0), "%"),
      expand = c(0.25, 0.25),
      name = "Percent (%)"
    )
} else if (input$y_axis_type == "rate") {
  i_chart <- i_chart +
    ggplot2::scale_y_continuous(
      labels = scales::label_percent(),     #     function(y) paste0(format(y, nsmall = 0), "%"),
      limits = c(0, 100),  # Fixed limit for rates (0-100%)
      expand = c(0, 0),    # Remove padding to keep limits exact
      name = "Percent (%)"
    )
} else if (input$y_axis_type == "currency") {
  i_chart <- i_chart +
    ggplot2::scale_y_continuous(
      labels = scales::dollar,
      expand = c(0.25, 0.25),
      name = "Amount (USD)"
    )
} else if (input$y_axis_type == "count") {
  i_chart <- i_chart +
    ggplot2::scale_y_continuous(
      labels = scales::comma,
      expand = c(0.25, 0.25),
      name = "Count"
    )
} else {
  i_chart <- i_chart +
    ggplot2::scale_y_continuous(
      labels = scales::number_format(accuracy = 0.01),
      expand = c(0.25, 0.25),
      name = "Value"
    )
}

  # Add fixed aspect ratio
  # i_chart <- i_chart + coord_fixed(ratio = 2/5)
  
  # Set base size for all future fonts
  i_chart <- i_chart + ggplot2::theme_minimal(
    base_size = 36,
    base_family = 'Cabin'
  ) 

  # Theme and title the plot object
  i_chart <- i_chart +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45,
                                          hjust = 1
      ),
      
      aspect.ratio = 2/5,    # This will create a 5:2 aspect ratio (width:height) WITHIN THE PANEL ONLY
      
      legend.position = "none",
      plot.title.position = "panel",
      text = ggplot2::element_text(color = "royalblue"),
      plot.title = ggtext::element_markdown(
        color = "darkgreen",
        size = ggplot2::rel(1.25),
        face = "bold",
        lineheight = 1.1,
        margin = ggplot2::margin(0, 0, 0, 0, "pt")
      ),
      plot.subtitle = ggtext::element_markdown(
        color = "darkgreen",
        size = ggplot2::rel(1.1),
        face = "bold",
        lineheight = 0.90,
        margin = ggplot2::margin(5, 0, 0, 0, unit = "pt")
      ),
      plot.caption = ggplot2::element_text(
        size = ggplot2::rel(.95),
        hjust = 0,
        vjust = 2,
        face = "italic",
        color = "darkblue"
      ),
      strip.text = ggplot2::element_text(
        color = "orange",
        size = ggplot2::rel(1.1),
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggplot2::element_text(color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA)
    )

  # Format and place text for Center Line
  i_chart <- i_chart +
    ggplot2::geom_text(
      aes(
        x = dplyr::first(!!sym(input$x_column)),
        y = dplyr::first(emp_cl),
        label = "Avg You Can Expect"
      ),
      color = "blue",
      vjust = -1,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place CL number
    ggplot2::geom_text(
      aes(
        x = dplyr::last(!!sym(input$x_column)),
        y = dplyr::last(emp_cl),
        label = format(round(dplyr::last(emp_cl), 2))
      ),
      color = "blue",
      vjust = 0,
      hjust = -0.1,
      size = 10
    )

  # Format and place text for emp UCL
  i_chart <- i_chart +
    ggplot2::geom_text(
      aes(
        x = dplyr::first(!!sym(input$x_column)) + 0.5,
        y = dplyr::first(emp_ucl),
        label = "Highest You Can Expect"
      ),
      color = "red",
      vjust = -1,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place number for emp_ucl
    ggplot2::geom_text(
      aes(
        x = dplyr::last(!!sym(input$x_column)),
        y = dplyr::last(emp_ucl),
        label = format(round(dplyr::last(emp_ucl), 2))
      ),
      color = "red",
      vjust = -1,
      hjust = -0.1,
      size = 10
    )

  # Format and place emp_lcl text and number
  i_chart <- i_chart +
    ggplot2::geom_text(
      aes(
        x = dplyr::first(!!sym(input$x_column)) + .5,
        y = dplyr::first(emp_lcl),
        label = "Lowest You Can Expect"
      ),
      color = "red",
      vjust = 1.75,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place number for emp_lcl
    ggplot2::geom_text(
      aes(
        x = dplyr::last(!!sym(input$x_column)),
        y = dplyr::last(emp_lcl),
        label = format(round(dplyr::last(emp_lcl), 2))
      ),
      color = "red",
      vjust = 1.5,
      hjust = -0.1,
      size = 10
    )

  # Add padding to individual plots
  i_chart <- i_chart +
    ggplot2::theme(
      plot.margin = ggplot2::margin(
        t = 0,
        r = 0,
        b = 0,
        l = 0
      )
    )

  # Save the plot to reactive values
  plot_store$untrended_i_chart <- i_chart
  
  # Return the chart
  return(i_chart)
})


############  END UNTRENDED I-CHART LOGIC  #####






################  BEGIN TRENDED I-CHART LOGIC  #####


output$trended_i_chart <- renderPlot({
  # Require necessary inputs
  req(
    filtered_data(), input$x_column, input$y_column, input$chart_title,
    input$year_range, input$multi_var_group, input$y_axis_type, input$divide_by_100
  )

  # Extract data from filtered_data()
  data <- filtered_data()

  # Make sure we have some data
  if (nrow(data) == 0) {
    return(ggplot() +
      annotate("text",
        x = 0.5,
        y = 0.5,
        label = "No data available with current filters"
      ) +
      theme_void())
  }

  # Process data for control limits calculation
  data_control_limits <- data

  # Process data according to divide_by_100 input
  if (input$divide_by_100 == "Yes") {
    # Divide values by 100 (e.g., convert 75 to 0.75)
    data_control_limits[[input$y_column]] <- data_control_limits[[input$y_column]] / 100
  }

  # For percentage display, convert to percentage for visualization
  if (input$y_axis_type == "Percentage") {
    data_control_limits[[input$y_column]] <- data_control_limits[[input$y_column]] * 100
  }

  ### CALCULATE TRENDED CL, SD AND CONTROL LIMITS ###
  # Fit a linear model to detect trend
  model_formula <- as.formula(paste(input$y_column, "~", input$x_column))
  model <- lm(model_formula, data = data_control_limits)
  model_summary <- summary(model)
  
  # Extract regression statistics
  p_value <- coef(model_summary)[2, 4]  # p-value for the slope
  r_squared <- model_summary$r.squared
  slope <- coef(model)[2]
  intercept <- coef(model)[1]
  
  # Calculate residual SD for control limits
  residuals_sd <- sd(residuals(model))
  
  # Create modified dataset for chart
  data_modified <- data
  
  # Process data according to divide_by_100 input
  if (input$divide_by_100 == "Yes") {
    # Divide values by 100 (e.g., convert 75 to 0.75)
    data_modified[[input$y_column]] <- data_modified[[input$y_column]] / 100
  }
  
  # For percentage display, convert to percentage for visualization
  if (input$y_axis_type == "Percentage") {
    data_modified[[input$y_column]] <- data_modified[[input$y_column]] * 100
  }
  
  # Calculate trended centerline and control limits
  data_modified$centerline <- predict(model, newdata = data_modified)
  data_modified$ucl <- data_modified$centerline + 3 * (residuals_sd/1.128)
  data_modified$lcl <- data_modified$centerline - 3 * (residuals_sd/1.128)
  
  # Set lower control limit to zero if negative (just like in untrended chart)
  data_modified$lcl <- pmax(data_modified$lcl, 0)
  
  # Add regression statistics to the data
  data_modified$p_value_clean <- p_value
  data_modified$r_sqr_value_clean <- r_squared
  data_modified$slope <- slope
  data_modified$intercept <- intercept
  
  # Shewhart Rule #1: violations if value is outside UCL or LCL
  sigma_signals <- data_modified[[input$y_column]] < data_modified$lcl | data_modified[[input$y_column]] > data_modified$ucl
  data_modified$emp_pts_out_of_control <- sigma_signals
  
  # Runs analysis for changing CL color and linetype
  runs <- sign(data_modified[[input$y_column]] - data_modified$centerline)
  runs <- runs[runs != 0]
  runs_lengths <- rle(runs)$lengths
  n_obs <- sum(runs_lengths)
  longest_run <- max(runs_lengths)
  n_runs <- length(runs_lengths)
  n_crossings <- n_runs - 1
  longest_run_max <- round(log2(n_obs) + 3)
  n_crossings_min <- qbinom(.05, n_obs - 1, 0.5)
  
  runs_signal <- (longest_run > longest_run_max | n_crossings < n_crossings_min)
  
  # Add runs_signal to data_modified
  data_modified <- data_modified %>%
    dplyr::mutate(
      runs_signal = as.logical(runs_signal) # Ensure it's a column, not a constant
    )
  
  # Replace zeros in the value column with NA
  data_modified[[input$y_column]] <- ifelse(data_modified[[input$y_column]] == 0,
    NA,
    data_modified[[input$y_column]]
  )
  
  # Compute x and y positions for geom_richtext
  x_pos <- min(data_modified[[input$x_column]], na.rm = TRUE)
  y_pos <- max(data_modified[[input$y_column]], na.rm = TRUE)
  
  # Calculate y-position for bottom annotations (15% below the minimum data point)
  bottom_y_pos <- min(data_modified[[input$y_column]], na.rm = TRUE) - 
    diff(range(data_modified[[input$y_column]], na.rm = TRUE)) * 0.15
  
  ###############  BEGIN TRENDED I-CHART TITLES  #####
  
  # Create a combined title from user-selected columns (same as in previous function)
  title_elements <- c()
  for(col_name in input$chart_title) {
    if(col_name %in% names(data_modified)) {
      if(length(unique(data_modified[[col_name]])) == 1) {
        title_elements <- c(title_elements, as.character(unique(data_modified[[col_name]])))
      } else if(length(unique(data_modified[[col_name]])) <= 5) {
        values <- paste(unique(data_modified[[col_name]]), collapse=", ")
        title_elements <- c(title_elements, values)
      } else {
        n_values <- length(unique(data_modified[[col_name]]))
        title_elements <- c(title_elements, paste0(n_values, " Values"))
      }
    }
  }
  
  # Combine title elements
  title_content <- if (length(title_elements) > 0) {
    paste(title_elements, collapse = " - ")
  } else {
    "Selected Data"
  }
  
  # Create the final title and subtitle
  trended_i_title <- paste0(
    "<span style='color:darkorchid1'>",
    "Expectation Chart (trended): ",
    title_content,
    "</span>"
  )
  
  # Modify this subtitle section in your line chart code
  req(input$year_range, input$staar_grouping)  # Make sure these inputs exist
  
  trended_i_subtitle <- paste0(
    "For Students Reaching <i>'",
    input$staar_grouping,
    "'</i><br>for Years ",  # Use HTML tags <i> for italics and <br> for line break
    input$year_range[1],
    " - ",
    input$year_range[2]
  )
  
  ###############  END TRENDED I-CHART TITLES  #####
  
  # Define y-axis LABEL based on User input
  y_axis_label <- input$y_column
  
  if (input$y_axis_type == "Percentage") {
    y_axis_label <- "Percentage (%)"
  } else if (input$y_axis_type == "Count") {
    y_axis_label <- "Count"
  } else if (input$y_axis_type == "Currency") {
    y_axis_label <- "Amount ($)"
  } else if (input$y_axis_type == "Rate") {
    y_axis_label <- "Percentage (%)"
  }
  
  # Create caption
  caption <- "\n Source: Texas Education Agency"
  
  # Compute the midpoint of the x-axis based on the year range
  x_midpoint <- mean(c(input$year_range[1], input$year_range[2]))
  
  # Create regression equation string
  reg_equation <- paste0(
    "y = ", 
    round(intercept, 2),
    ifelse(slope >= 0, " + ", " - "),
    round(abs(slope), 2),
    "x"
  )
  
  # Create plot object TRENDED I-chart USING GGPLOT2 ONLY
  trended_i_chart <- ggplot2::ggplot(
    data_modified,
    aes(
      x = .data[[input$x_column]],
      y = .data[[input$y_column]]
    )
  ) +
    
    # Add lines connecting the points
    ggplot2::geom_line(
      color = "darkgray",
      linewidth = 2.0
    ) +
    
    # Use 'color' to map logical TRUE/FALSE to red/black points
    ggplot2::geom_point(aes(color = factor(emp_pts_out_of_control)),
      size = 8
    ) +
    
    ggplot2::scale_color_manual(values = c(
      "TRUE" = "red",
      "FALSE" = "blue"
    )) +
    
    # Trended centerline - line type depends on value of runs_signals
    geom_line(aes(
      y = centerline,
      linetype = factor(runs_signal)
    ),
    color = "blue") +
    
    scale_linetype_manual(values = c(
      "FALSE" = "solid",
      "TRUE" = "dashed"
    )) +
    
    # Format ucl LINE
    ggplot2::geom_line(aes(y = ucl),
      color = "red",
      linetype = "solid",
      linewidth = 2
    ) +
    
    # Format lcl LINE
    ggplot2::geom_line(aes(y = lcl),
      color = "red",
      linetype = "solid",
      linewidth = 2
    ) +
    
    # Add labels
    ggplot2::labs(
      title = trended_i_title,
      subtitle = trended_i_subtitle,
      caption = caption,
      x = element_blank(), 
      y = element_blank()
    ) +
    
    # Disclose CL calculation - centered on x-axis
    ggtext::geom_richtext(
      aes(
        label = paste0("<b><i>Highest, Lowest and Avg Expected Values",
        "<br>",
        "Account for Trend in Your Process</i></b>"),
        x = x_midpoint,  # Use midpoint of x values to center box on x axis
        y = y_pos
      ),
      size = 10,
      color = "black",
      hjust = .50,  # Set to 0 for left justification, .5 for center
      vjust = -1,
      fill = "lightblue",
      label.color = "black",
      label.size = unit(0.5, "inch"),  # Make the text outline box larger
      label.padding = unit(c(10, 10, 5, 10), "pt"),  # top, right, bottom, left padding
      lineheight = 0.8  # Reduce the line height to bring lines closer together
    )
  
  # Configure x-axis to show all years
  if (is.numeric(data_modified[[input$x_column]])) {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_x_continuous(
        breaks = unique(data_modified[[input$x_column]]),
        expand = c(
          0.15,
          0.00
        )
      )
  } else {
    # For character/factor columns, use discrete scale
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_x_discrete(
        expand = c(
          0.15,
          0.00
        )
      )
  }
  
  # Configure y-axis based on chart type with extended lower range for text boxes
  if (input$y_axis_type == "percent") {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_y_continuous(
        labels = function(y) paste0(format(y, nsmall = 0), "%"),
        expand = c(0.25, 0.25),
        name = "Percent (%)",
        # Add extra space at the bottom for text annotations
        limits = function(y) c(min(y) - diff(range(y))*0.20, max(y) + diff(range(y))*0.05)
      )
  } else if (input$y_axis_type == "rate") {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_y_continuous(
        labels = scales::label_percent(),
        limits = c(0, 100),  # Fixed limit for rates (0-100%)
        expand = c(0, 0.20),    # Add padding at bottom for labels
        name = "Percent (%)"
      )
  } else if (input$y_axis_type == "currency") {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_y_continuous(
        labels = scales::dollar,
        expand = c(0.25, 0.25),
        name = "Amount (USD)",
        # Add extra space at the bottom for text annotations
        limits = function(y) c(min(y) - diff(range(y))*0.20, max(y) + diff(range(y))*0.05)
      )
  } else if (input$y_axis_type == "count") {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_y_continuous(
        labels = scales::comma,
        expand = c(0.25, 0.25),
        name = "Count",
        # Add extra space at the bottom for text annotations
        limits = function(y) c(min(y) - diff(range(y))*0.20, max(y) + diff(range(y))*0.05)
      )
  } else {
    trended_i_chart <- trended_i_chart +
      ggplot2::scale_y_continuous(
        labels = scales::number_format(accuracy = 0.01),
        expand = c(0.25, 0.25),
        name = "Value",
        # Add extra space at the bottom for text annotations
        limits = function(y) c(min(y) - diff(range(y))*0.20, max(y) + diff(range(y))*0.05)
      )
  }
  
  # Set base size for all future fonts
  trended_i_chart <- trended_i_chart + ggplot2::theme_minimal(
    base_size = 36,
    base_family = 'Cabin'
  ) 
  
  # Theme and title the plot object
  trended_i_chart <- trended_i_chart +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45,
                                          hjust = 1
      ),
      
      aspect.ratio = 2/5,    # This will create a 5:2 aspect ratio (width:height) WITHIN THE PANEL ONLY
      
      legend.position = "none",
      plot.title.position = "panel",
      text = ggplot2::element_text(color = "royalblue"),
      plot.title = ggtext::element_markdown(
        color = "darkgreen",
        size = ggplot2::rel(1.25),
        face = "bold",
        lineheight = 1.1,
        margin = ggplot2::margin(0, 0, 0, 0, "pt")
      ),
      plot.subtitle = ggtext::element_markdown(
        color = "darkgreen",
        size = ggplot2::rel(1.1),
        face = "bold",
        lineheight = 0.90,
        margin = ggplot2::margin(5, 0, 0, 0, unit = "pt")
      ),
      plot.caption = ggplot2::element_text(
        size = ggplot2::rel(.95),
        hjust = 0,
        vjust = 2,
        face = "italic",
        color = "darkblue"
      ),
      strip.text = ggplot2::element_text(
        color = "orange",
        size = ggplot2::rel(1.1),
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggplot2::element_text(color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA)
    )
  
  # Get first and last values for trend lines to display text
  first_x <- dplyr::first(data_modified[[input$x_column]])
  last_x <- dplyr::last(data_modified[[input$x_column]])
  
  # Format and place text for Center Line - adapted for trended
  trended_i_chart <- trended_i_chart +
    ggplot2::geom_text(
      aes(
        x = first_x,
        y = dplyr::first(centerline),
        label = "Avg You Can Expect"
      ),
      color = "blue",
      vjust = -1,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place CL number
    ggplot2::geom_text(
      aes(
        x = last_x,
        y = dplyr::last(centerline),
        label = format(round(dplyr::last(centerline), 2))
      ),
      color = "blue",
      vjust = 0,
      hjust = -0.1,
      size = 10
    )
  
  # Format and place text for UCL
  trended_i_chart <- trended_i_chart +
    ggplot2::geom_text(
      aes(
        x = first_x + 0.5,
        y = dplyr::first(ucl),
        label = "Highest You Can Expect"
      ),
      color = "red",
      vjust = -1,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place number for ucl
    ggplot2::geom_text(
      aes(
        x = last_x,
        y = dplyr::last(ucl),
        label = format(round(dplyr::last(ucl), 2))
      ),
      color = "red",
      vjust = -1,
      hjust = -0.1,
      size = 10
    )
  
  # Format and place lcl text and number
  trended_i_chart <- trended_i_chart +
    ggplot2::geom_text(
      aes(
        x = first_x + .5,
        y = dplyr::first(lcl),
        label = "Lowest You Can Expect"
      ),
      color = "red",
      vjust = 1.75,
      hjust = 0.50,
      size = 10
    ) +
    # Format and place number for lcl
    ggplot2::geom_text(
      aes(
        x = last_x,
        y = dplyr::last(lcl),
        label = format(round(dplyr::last(lcl), 2))
      ),
      color = "red",
      vjust = 1.5,
      hjust = -0.1,
      size = 10
    )
  
  # Add statistical information at bottom of plot
  trended_i_chart <- trended_i_chart +
    # Add p-value text at bottom left
    ggtext::geom_richtext(
      aes(
        x = min(data_modified[[input$x_column]], na.rm = TRUE),
        y = bottom_y_pos
      ),
      label = paste("*p*-value:", round(p_value, 3)),
      fill = "lightblue",
      label.color = "black",
      size = 12,
      hjust = 0,
      vjust = 6.0
    ) +
    
    # Add R^2 value text at bottom right
    ggtext::geom_richtext(
      aes(
        x = max(data_modified[[input$x_column]], na.rm = TRUE),
        y = bottom_y_pos
      ),
      label = paste("Unadj *R*<sup>2</sup> =", round(r_squared, 2)),
      fill = "lightgreen",
      label.color = "black",
      size = 12,
      hjust = 1,
      vjust = 6.0
    ) +
    
    # Add regression equation text in middle bottom
    ggtext::geom_richtext(
      aes(
        x = x_midpoint,
        y = bottom_y_pos
      ),
      label = paste("<b>Equation:</b>", reg_equation),
      fill = "lightyellow",
      label.color = "black",
      size = 12,
      hjust = 0.5,
      vjust = 6.0
    )
  
  # Add padding to individual plots
  trended_i_chart <- trended_i_chart +
    ggplot2::theme(
      plot.margin = ggplot2::margin(
        t = 0,
        r = 0,
        b = 0,
        l = 0
      )
    )
  
  # Save the plot to reactive values
  plot_store$trended_i_chart <- trended_i_chart
  
  # Return the chart
  return(trended_i_chart)
})


############  END TRENDED I-CHART LOGIC  #####



####################  BEGIN COHORT CHART LOGIC  #####

# Print debugging info
  print("Rendering cohort chart")

output$cohort_chart <- renderPlot({
  
  # Require data
  req(cohort_data())
  data <- cohort_data()
  
  print(paste("Cohort data has", nrow(data), "rows"))
  
  # Make sure we have data
  if (nrow(data) == 0) {
    print("No cohort data available")
    return(ggplot() +
      annotate("text", x = 0.5, y = 0.5, 
               label = "No cohort data available with current selections.\nTry adjusting filters or cohort settings.") +
      theme_void())
  }
  
  # Create title elements ONLY from the current cohort data
  title_elements <- c()
  for(col_name in input$chart_title) {
    if(col_name %in% names(data)) {
      # Use ONLY values from the current cohort data
      cohort_values <- unique(data[[col_name]])
      
      if(length(cohort_values) == 1) {
        title_elements <- c(title_elements, as.character(cohort_values))
      } else if(length(cohort_values) <= 5) {
        values <- paste(cohort_values, collapse=", ")
        title_elements <- c(title_elements, values)
      } else {
        n_values <- length(cohort_values)
        title_elements <- c(title_elements, paste0(n_values, " Values"))
      }
    }
  }
  
  # Combine title elements
  title_content <- if (length(title_elements) > 0) {
    paste(title_elements, collapse = " - ")
  } else {
    "Selected Data"
  }
  
  # Create cohort-specific title
  title <- paste0(
    "<span style='color:darkorchid1'>",
    "Cohort Progression: ", 
    title_content,
    "</span>"
  )
  
  subtitle <- paste0(
    "Tracking cohort from Grade ", input$cohort_start_grade, 
    " (", input$cohort_start_year, ")"
  )
  
  # Create y-axis label based on type
  y_axis_label <- input$y_column
  
  if (input$y_axis_type == "Percentage") {
    y_axis_label <- "Percentage (%)"
  }  else if (input$y_axis_type == "Rate") {
    y_axis_label <- "Percentage (%) Count"
  }  else if (input$y_axis_type == "Count") {
    y_axis_label <- "Count"
  } else if (input$y_axis_type == "Currency") {
    y_axis_label <- "Amount ($)"
  }
  
  # Create the chart using progression as x-axis
  p <- ggplot(data,
              aes(x = progression,
                  y = .data[[input$y_column]])) +
    geom_line(group = 1, color = "darkgray", linewidth = 2.0) +
    geom_point(size = 8, color = "blue") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = "\n Source: Texas Education Agency",
      x = element_blank(),
      y = element_blank()
      # x = "Academic Progression",
      # y = y_axis_label
    ) +
    theme_minimal(
      base_size = 36,
      base_family = 'Cabin'
      ) +
    
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_line(color = "gray", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      plot.title.position = "panel",
      text = element_text(color = "royalblue"),
      plot.title = ggtext::element_markdown(
        color = "darkgreen",
        size = rel(1.25),
        face = "bold",
        lineheight = 1.1,
        margin = margin(0, 0, 0, 0, "pt")
      ),
      plot.subtitle = ggtext::element_markdown(
        color = "darkgreen",
        size = rel(1.1),
        face = "bold",
        lineheight = 0.90,
        margin = margin(5, 0, 0, 0, "pt")
      ),
      plot.caption = element_text(
        size = rel(.95),
        hjust = 0,
        vjust = 2,
        face = "italic",
        color = "darkblue"
      ),
      axis.text = element_text(color = "black")
    )
  
  # Configure y-axis based on chart type
  if (input$y_axis_type == "Percentage") {
    p <- p + scale_y_continuous(
      labels = function(y) paste0(format(y, nsmall = 0), "%"),
      expand = c(0.25, 0.25)
    )
  } else if (input$y_axis_type == "Currency") {
    p <- p + scale_y_continuous(
      labels = scales::dollar,
      expand = c(0.25, 0.25)
    )
  } else {
    p <- p + scale_y_continuous(
      labels = scales::comma,
      expand = c(0.25, 0.25)
    )
  }
  
  # Save the plot for download
  plot_store$cohort_chart <- p
  
  # Return the plot
  return(p)
  
})


####################  END COHORT CHART LOGIC  #####



####################  BEGIN DOWNLOAD HANDLERS  #####


# Create a reactive values object to store the latest plot objects
plot_store <- reactiveValues(
  run_chart = NULL,
  line_chart = NULL,
  bar_chart = NULL,
  untrended_i_chart = NULL,
  trended_i_chart = NULL,
  cohort_chart = NULL  # Add this line
)



# Download handlers for all charts - PNG only

# Add download handler for run chart

output$download_run_chart <- downloadHandler(
  filename = function() {
    paste("run-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      # Check if plot exists
      if (is.null(plot_store$run_chart)) {
        message("Run chart is NULL. Please create a chart first.")
        return()
      }
      
      # PNG export only
      ggsave(file, plot = plot_store$run_chart, device = "png", 
             width = 12, height = 8, dpi = 300,
             limitsize = FALSE)
    }, error = function(e) {
      # Log the error
      message("Error in download_run_chart: ", e$message)
    })
  }
)



# Add download handler for line chart

output$download_line_chart <- downloadHandler(
  filename = function() {
    paste("line-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      # Check if plot exists
      if (is.null(plot_store$line_chart)) {
        message("Line chart is NULL. Please create a chart first.")
        return()
      }
      
      # PNG export only
      ggsave(file, plot = plot_store$line_chart, device = "png", 
             width = 12, height = 8, dpi = 300,
             limitsize = FALSE)
    }, error = function(e) {
      # Log the error
      message("Error in download_line_chart: ", e$message)
    })
  }
)



# Add download handler for bar chart

output$download_bar_chart <- downloadHandler(
  filename = function() {
    paste("bar-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      # Check if plot exists
      if (is.null(plot_store$bar_chart)) {
        message("Bar chart is NULL. Please create a chart first.")
        return()
      }
      
      # PNG export only
      ggsave(file, plot = plot_store$bar_chart, device = "png", 
             width = 12, height = 8, dpi = 300,
             limitsize = FALSE)
    }, error = function(e) {
      # Log the error
      message("Error in download_bar_chart: ", e$message)
    })
  }
)




# Add download handler for untrended I chart

output$download_untrended_i_chart <- downloadHandler(
  filename = function() {
    paste("untrended-i-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      # Check if plot exists
      if (is.null(plot_store$untrended_i_chart)) {
        message("I-chart is NULL. Please create a chart first.")
        return()
      }
      
      # PNG export only
      ggsave(file, plot = plot_store$untrended_i_chart, device = "png", 
             width = 14, height = 10, dpi = 300,
             limitsize = FALSE)
    }, error = function(e) {
      # Log the error
      message("Error in download_untrended_i_chart: ", e$message)
    })
  }
)



# Add download handler for trended I chart

output$download_trended_i_chart <- downloadHandler(
  filename = function() {
    paste("trended-i-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      # Check if plot exists
      if (is.null(plot_store$trended_i_chart)) {
        message("I-chart is NULL. Please create a chart first.")
        return()
      }
      
      # PNG export only
      ggsave(file, plot = plot_store$trended_i_chart, device = "png", 
             width = 14, height = 10, dpi = 300,
             limitsize = FALSE)
    }, error = function(e) {
      # Log the error
      message("Error in download_trended_i_chart: ", e$message)
    })
  }
)



# Add download handler for cohort chart

output$download_cohort_chart <- downloadHandler(
  filename = function() {
    paste("cohort-chart-", format(Sys.Date(), "%Y%m%d"), ".png", sep = "")
  },
  content = function(file) {
    tryCatch({
      if (is.null(plot_store$cohort_chart)) {
        message("Cohort chart is NULL. Please create a chart first.")
        return()
      }
      
      # Add explicit white background to the plot before saving
      # Also add more generous margins and increase the size of axis elements
      p <- plot_store$cohort_chart +
        theme(
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "white", color = NA),
          plot.margin = margin(0.5, 1.5, 0.5, 0.5, "in"),  # top, right, bottom, left
          
          # Increase size of axis text elements
          axis.text.x = element_text(size = rel(1.5), face = "bold"),
          axis.text.y = element_text(size = rel(1.5), face = "bold"),
          
          # Increase size of axis titles
          axis.title.x = element_text(size = rel(1.5), face = "bold", margin = margin(t = 20)),
          axis.title.y = element_text(size = rel(1.5), face = "bold", margin = margin(r = 20)),
          
          # Ensure tick marks are more visible
          axis.ticks = element_line(size = 1),
          axis.ticks.length = unit(0.15, "cm")
        )
      
      # Use wider dimensions and ensure titles fit
      ggsave(file, plot = p, device = "png", 
             width = 14, height = 9, dpi = 300,
             limitsize = FALSE, bg = "white")
    }, error = function(e) {
      message("Error in download_cohort_chart: ", e$message)
    })
  }
)

####################  END DOWNLOAD HANDLERS  #####




############  END OF SERVER LOGIC SECTION  #####


```




<!-- ####################  BEGIN VISUALIZATIONS SECTION -->


# Data Table {data-icon="fa-signal" data-width=650}

```{r display-table}

# Output Table title with styling
tags$h2("CLICK ON COLUMNS TO SORT",
        style = "text-align: center;
        font-size: 20px;
        font-weight: bold;
        color: #2E86C1;")


DTOutput("data_frame_chosen")

```



# Run Chart  {data-icon="fa-signal" data-width=200 data-height=200}

```{r display-run-chart}

div(class = "download-section",
    downloadButton("download_run_chart", "Download Chart")
)

plotOutput("run_chart_plot")

```




# Bar Chart  {data-icon="fa-signal" data-width=650}

```{r display-bar-chart}
div(class = "download-section",
    downloadButton("download_bar_chart", "Download Chart")
)

plotOutput("bar_chart")

```


# Line Chart  {data-icon="fa-signal" data-width=650}

```{r display-line-chart}
div(class = "download-section",
    downloadButton("download_line_chart", "Download Chart")
)

plotOutput("line_chart",
           height = "800px")

```


# Untrended I-Chart  {data-icon="fa-signal" data-width=650}

```{r display-untrended-i-chart}
div(class = "download-section",
    downloadButton("download_untrended_i_chart", "Download Chart")
)

plotOutput("untrended_i_chart", height = "800px")

```



# Trended I-Chart  {data-icon="fa-signal" data-width=650}

```{r display-trended-i-chart}
div(class = "download-section",
    downloadButton("download_trended_i_chart", "Download Chart")
)

plotOutput("trended_i_chart", height = "800px")

```



# Cohort Progression {data-icon="fa-users" data-width=650}

```{r display-cohort-chart}
div(class = "download-section",
    downloadButton("download_cohort_chart", "Download Chart")
)

plotOutput("cohort_chart", height = "600px")

```

# Debug Info
```{r}
# Debug Info {data-height=200}
verbatimTextOutput("debug_info")
```



<!-- ####################  END VISUALIZATIONS SECTION -->




```{r, cleaning-survey-data, eval=FALSE}
# 
# library(readr)
# library(tidyverse)
# library(here)
# 
# 
# # Read the original CSV
# raw_data <- read_csv(here(
# "data", "SCUC Teacher-Student Relationships Surveys.csv")
# )
# 
# 
# # Convert percentage column to numeric
# raw_data$`Percent Favorable` <- as.numeric(gsub("[%]", "", raw_data$`Percent Favorable`)) / 100
# raw_data$`Sense of Belonging` <- as.numeric(gsub("[%]", "", raw_data$`Sense of Belonging`)) / 100
# 
# glimpse(raw_data)
# 
# # Write the cleaned data to a new CSV
# write_csv(raw_data, "data/cleaned_SCUC_Teacher-Student Relationships Surveys.csv")

```


```{r, removing-problems-from-survey-data, eval=FALSE}

# 
# library(readr)
# library(tidyverse)
# library(here)
# 
# 
# 
# # Read the problematic file
# problem_data <- read_csv(here("data", "cleaned_SCUC_Teacher-Student Relationships Surveys.csv")
# )
# 
# # Remove rows where Level is NA or "Video about survey results:"
# cleaned_data <- problem_data %>%
#   filter(!is.na(Level) & Level != "Video about survey results:")
# 
# 
# # Optional: Remove rows where both Percent Favorable and Sense of Belonging are NA
# cleaned_data <- cleaned_data %>%
#   filter(!(is.na(`Percent Favorable`) & is.na(`Sense of Belonging`)))
# 
# 
# # Write the cleaned file
# write_csv(cleaned_data, "data/cleaned_SCUC_Teacher-Student Relationships Surveys.csv")

```



