---
title: "How to Build Titles and Subtitles"
author: "Dan Swart"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

### Here's a step-by-step explanation of how the title and subtitle are created and displayed in this advanced line plot code.  It assumes the plotting code utilizes a 'group' argument to create the plot.
<br>

The dataframe chosen by the User is monitored and updated by the Server.

```{r, eval=FALSE}

# File upload chooser button
fileInput(inputId = "file_chosen",
          label = "Choose a CVS File",
          accept = c(".csv")) 


# File upload observer feeds 'column_data' to global data storage called 'column_data'
observeEvent(input$file_chosen, {
  req(input$file_chosen)

  # Read the CSV file
  data <- read_csv(input$file_chosen$datapath, col_names = TRUE)
  
  # Store data frame selected by user.  Updated when different data frame is chosen by user  
  data_reactive(data)     

  # Store column names in imported data frame.  Updated when different data frame is chosen by user
  column_choices(names(data))
  
  # Store unique values of row contents of each column name.  Updated when different data frame is chosen by user
  column_data(lapply(data, unique))     
  
})   

```

<br>

#### In the global r-chunk, 'column_choices' stores and updates column names available for selection during the runtime of the app.


```{r, eval=FALSE}

column_choices <- reactiveVal(NULL) 

```


#### The Server monitors 'column_choices', creates inputID 'chart_title' and adds 'output$chart_title_selected' to the OUTPUT dataframe storing the title names selected by the User.


```{r, eval=FALSE}

# Generate dropdown of column names for chart title selection
output$chart_title_selected <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "chart_title",
    label = "Select Columns for Chart Title:",
    choices = column_choices(),
    selected = column_choices()[1],  # Default to first column
    multiple = TRUE                  # Allow selecting multiple columns
  )
})


# inputID = "chart_title" populates this
# dropdown for Chart Title selection
uiOutput("chart_title_selected")      

```

#### The Server also monitors 'column_choices', creates inputID 'chart_subtitle' and adds 'output$chart_sutitle_selected' to the OUTPUT dataframe storing the subtitle names selected by the User.


```{r, eval=FALSE}
}
# Generate dropdown of column names for chart subtitle selection
output$chart_subtitle_selected <- renderUI({
  req(column_choices())
  selectInput(
    inputId = "chart_subtitle",
    label = "Select Value for Chart Subtitle:",
    choices = column_choices(),                  # these are row names of selected columns
    selected = "year"                            # column_choices()[2] # Default selection
  )
})

# inputID = "chart_subtitle" populates this
# dropdown for Chart Subtitle column selection
uiOutput("chart_subtitle_selected")   

```


#### The Server also monitors 'column_choices', creates inputID 'advanced_line_group' and adds 'output$advanced_line_group_selected' to the OUTPUT dataframe storing the names selected by the User.


```{r, eval=FALSE}

# ADVANCED multi-line chart dropdown (for multiple category selection)
output$advanced_line_group_selected <- renderUI({
  req(column_choices())
  selectizeInput(  # Using selectizeInput for better multiple selection
    inputId = "advanced_line_group",
    label = "For Line Charts:  Select Categories for Detailed Comparison:",
    choices = column_choices(),
    selected = "group",
    multiple = TRUE,  # Allow multiple selections
    options = list(plugins = list('remove_button'))
  )
})


# inputID = "advanced_line_group" populates this
# dropdown for selecting grouping variable(s) for all charts that can display more than 1 variable
uiOutput("advanced_line_group_selected")



```

<br>
#### Store the names collected by the 'advanced_line_group' dropdown in the object 'groups_title'


```{r, eval=FALSE}

# Create a string containing all title names selected
groups_title <- paste(input$advanced_line_group, collapse = ", ")

```

<br>
This creates a string showing which columns are being used for grouping (from input$advanced_line_group), joining them with commas.

<br>
#### Build Title Elements


```{r, eval=FALSE}

# Examines length of string in input$chart_title and prepares the 'title_elements' object based on length of string
title_elements <- c()                    # start with an empty vector
for(col_name in input$chart_title) {     # loop thru each column name in 'chart_title'
  if(col_name %in% names(data)) {        # if column name is within the names found in data frame
    if(length(unique(data[[col_name]])) == 1) {     # if the string has length of one
      title_elements <- c(title_elements, as.character(unique(data[[col_name]])))     # put that one name in the object 'title_elements'
    } else if(length(unique(data[[col_name]])) <= 5) {     # or, if the length is < 5, put the names into the object 'title_elements'
      values <- paste(unique(data[[col_name]]), collapse=", ")
      title_elements <- c(title_elements, values)
    } else {     # if there are more than 5 names, just pint 'x number of values'
      n_values <- length(unique(data[[col_name]]))
      title_elements <- c(title_elements, paste0(n_values, " Values"))
    }
  }
}

```

<br>
This code:

* Starts with an empty vector named title_elements
* Loops through each column selected in input$chart_title
* Adds title content based on the data in each column:

    - If there's only 1 unique value: uses that value
    - If there are 2-5 unique values: lists all values separated by commas
    - If there are more than 5 values: just shows the count (e.g., "10 Values")


<br>
#### Create the Final Title Content object


```{r, eval=FALSE}

# Combine title elements if there is at least one value in the 'title_elements' string
title_content <- if (length(title_elements) > 0) {
  paste(title_elements, collapse = " - ")
} else {
  "Selected Data"
}

```

<br>
'title_content' now combines all collected title_elements, puts dashes between them, or uses "Selected Data" if nothing was collected.


<br>
#### Format Title and Subtitle with HTML/CSS 

The entire formatted Title Elements are now stored in 'title_content'.'  (the Subtitle does not use the contents of 'title_content'

```{r, eval=FALSE}

# Create the final title and subtitle
title <- paste0(
  "<span style='color:darkorchid1'>",
  "Multi-Line Chart: ", 
  title_content,
  "</span>"
)

subtitle <- paste0(
  "for Years ", input$year_range[1], " - ", input$year_range[2]
  # " (Comparing by: ", groups_title, ")"
)

```


* The title gets HTML styling with darkorchid1 color and includes the combined title elements
* The subtitle shows the year range and what groups are being compared

<br>

#### Apply Title and Subtitle to the Plot

```{r, eval=FALSE}

labs(
  title = title,
  subtitle = subtitle,
  caption = paste(" Comparison by ", groups_title),
  x = input$x_column,
  y = input$y_column,
  color = "Categories"  # Legend title
)

```

<br>
* The formatted strings are passed to the labs() function which sets them as the plot's title and subtitle.


#### Style the Title and Subtitle

```{r, eval=FALSE}

theme(
  plot.title.position = "panel",
  plot.title = ggtext::element_markdown(
     color = "darkgreen",
     size = rel(1.25),
     face = "bold",
     lineheight = 1.3,
     margin = margin(0, 0, 0.25, 0, "lines")
  ),
  
  plot.subtitle = ggtext::element_markdown(
     color = "darkgreen",
     size = rel(1.10),
     face = "bold",
     lineheight = 1.3,
     margin = margin(0, 0, 0.5, 0, "lines")
  ),
 )

```

<br>
This applies specific styling to the title and subtitle:

* Uses element_markdown() to render HTML/CSS in the title
* Sets both to dark green color
* Makes the title 25% larger than base font
* Makes the subtitle 10% larger than base font
* Both are bold with specific line spacing and margins










